### 2021-07-05 rad-icfp-2021/intro.tex
Part1: Subscripts and math environments in listings

\begin{example}[First-order term]
On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

@@

    \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1,y2,y3,y4,y5 = 0,0,0,0,1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
This term can be further optimised via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These return all the intermediate result of a fold in an array.
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

>>>>

\begin{example}[First-order term]
On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$, ..., y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$, ..., y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
        !w$_2'$(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

@@

    \begin{center}
            \begin{tabular}{l}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !let y$_1$,y$_2$,y$_3$,y$_4$,y$_5$ = 0,0,0,0,1 in!\\
                !let y$_1'$! != y$_1$+w$_1$*y$_5$ in!\\
                !let y$_4'$! != y$_4$+x$_1$*y$_5$ in!\\
                !(y$_1'$+x$_2$*y$_4'$!!, y$_2$+x$_1$*y$_4'$!!, y$_3$)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
This term can be further optimized via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !(w$_1$+x$_2$*x$_1$, x$_1$*x$_1$, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !shift1 [v$_1$,...,v$_n$] = [1,v$_1$,...,v$_{n-1}$]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These return all the intermediate result of a fold in an array.
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

### 2021-07-05 rad-icfp-2021/purerad.tex
Part1: Subscripts and math environments in listings

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
        =& !let w1,Y1=!\\
        & \quad\quad !let y11,Y11= <x1, fun (y1,y2,y3,z) -> Y(y1+z,y2,y3)> in! \\
        & \quad\quad !let y12,Y12= <x2, fun (y1,y2,y3,y4,z) -> Y11(y1,y2+z,y3,y4)> in! \\
        & \quad\quad !<y11 * y12, fun (y1,y2,y3,z) -> Y12(y1,y2,y3,y12*z,y11*z) > in! \\
        & !let w2,Y2=!\\
        & \quad\quad !let y21,Y21= <w1, fun (y1,y2,y3,y4,z) -> Y(y1,y2,y3,y4+z)> in! \\
        & \quad\quad !let y22,Y22= <x1, fun (y1,y2,y3,y4,y5,z) -> Y11(y1+z,y2,y3,y4,y5)> in! \\
        & \quad\quad !<y21 * y22, fun (y1,y2,y3,y4,z) -> Y22(y1,y2,y3,y4,y22*z,y21*z) > in! \\
        & !let y,Y3= <w, fun (y1,y2,y3,y4,z) -> Y2(y1,y2,y3,y4+z)> in! \\
        & !<y, fun (y1,y2,y3,z) -> Y3(y1,y2,y3,0,z) >!
    \end{tabular}
    \medskip

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
        =& !let x,Y1= <0, fun (X,z) -> Y(X)> in! \\
        & !let B,Y2= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
        & !let A0= scanl * x B in!\\
        & !let A1= map2 (a,b.b) A0 B in!\\
        & !let A2= shift1 (map2 (a,b.a) A0 B) in!\\
        & !let A3= scanr * 1 A1 in!\\
        & !<prod(B), fun (X,z) -> Y2(X,0,map2 (a,b. a*z*b) A2 A3)>! 
    \end{tabular}
\end{example}

The idea is that $\rho$ represents the return type of the derivative part, which should be !A1x...,xAn! 
if we want the whole gradient of a term !e! in context $\Gamma$=!x1:A1,...,xn:An!.
The subscript $\Gamma$ represents the current context, 
which is locally augmented for instance when differentiating a !let! rule.
For non-unary operations, we differentiate the arguents from left to right and add their derivatives to the current stack, 
which is modeled by the continuation $Y$. 
Importantly for performance, each continuation variable $Y$ is only used once.

@@

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x1,...,xn) -> (fun (y1,...,yn) -> (f1,...,fn))(e1,...,en)! 
$\transto$ 
!fun (x1,...,xn) -> let y1,...,yn = e1,...,en in (f1,...,fn)!
$\transto$
!fun (x1,...,xn) -> (f1[e1/y1],...,fn[en/yn])!

@@

    \begin{tabular}{c l}
        & $\grad_\Gamma$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
        =& TODO \\
    \end{tabular}
    \medskip

>>>>

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$!) \\
        =& !let w$_1$,Y$_1$=!\\
        & \quad\quad !let y$_1$1,Y$_1$1= <x$_1$, fun (y$_1$,y$_2$,y$_3$,z) -> Y(y$_1$+z,y$_2$,y$_3$)> in! \\
        & \quad\quad !let y$_{12}$,Y$_{12}$= <x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{11}$(y$_{1}$,y$_{2}$+z,y$_{3}$,y$_{4}$)> in! \\
        & \quad\quad !<y$_{11}$ * y$_{12}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{12}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{12}$*z,y$_{11}$*z) > in! \\
        & !let w$_{2}$,Y$_{2}$=!\\
        & \quad\quad !let y$_{21}$,Y$_{21}$= <w$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
        & \quad\quad !let y$_{22}$,Y$_{22}$= <x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$,z) -> Y$_{11}$(y$_{1}$+z,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$)> in! \\
        & \quad\quad !<y$_{21}$ * y$_{22}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{22}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{22}$*z,y$_{21}$*z) > in! \\
        & !let y,Y$_{3}$= <w, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
    \end{tabular}
    \medskip

    \begin{tabular}{c l}
        &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
        =& !let x,Y$_{1}$= <0, fun (X,z) -> Y(X)> in! \\
        & !let B,Y$_{2}$= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
        & !let A$_{0}$= scanl * x B in!\\
        & !let A$_{1}$= map2 (a,b.b) A$_{0}$ B in!\\
        & !let A$_{2}$= shift1 (map2 (a,b.a) A$_{0}$ B) in!\\
        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
        & !<prod(B), fun (X,z) -> Y$_{2}$(X,0,map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
    \end{tabular}
\end{example}

The idea is that $\rho$ represents the return type of the derivative part, which should be !A$_{1} \times \ldots \times$ A$_n$! 
if we want the whole gradient of a term !e! in context !$\Gamma = \;$x$_{1}$:A$_{1}$,...,x$_n$:A$_n$!.
The subscript $\Gamma$ denotes the current context, 
which is locally augmented for instance when differentiating a !let! rule.
For non-unary operations, we differentiate the arguments from left to right and add their derivatives to the current stack, 
which is modeled by the continuation $Y$. 
Importantly for performance, each continuation variable $Y$ is only used once.

@@

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x$_{1}$,...,x$_n$) -> (fun (y$_{1}$,...,yn) -> (f$_{1}$,...,f$_n$))(e$_{1}$,...,e$_n$)! 
$\transto$ 
!fun (x$_{1}$,...,x$_n$) -> let y$_{1}$,...,y$_n$ = e$_1$,...,e$_n$ in (f$_{1}$,...,f$_n$)!
$\transto$
!fun (x$_{1}$,...,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],...,f$_n$[e$_n$/y$_n$])!

@@

    \begin{tabular}{c l}
        & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
        =& TODO \\
    \end{tabular}
    \medskip
